---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# MoNAn

<!-- badges: start -->
<!-- badges: end -->

MoNAn is the software implementation of the statistical model outlined in:

Block, P., Stadtfeld, C., & Robins, G. (2022). A statistical model for the analysis of mobility 
tables as weighted networks with an application to faculty hiring networks. Social Networks, 68, 264-278.

[Link to the study in Social Networks](https://www.sciencedirect.com/science/article/pii/S0378873321000654).

[Pre-print with minor differences to the journal version](https://osf.io/preprints/socarxiv/n86rx/).

The model allows for the analysis of emergent structures (known from network analysis) in mobility tables, alongside exogenous predictors. It combined features from classical log-linear models for mobility tables and Exponential Random Graph Models (ERGMs). In the absence of emergent structures, the models reduces to the former, when ignoring characteristics of individuals, it is an ERGM for weighted networks.

Announcements about workshops etc. can be found [here](https://www.suz.uzh.ch/de/institut/professuren/block/monan_software.html).


<span style="color: red;">
# Warning!!!

From Sat. 08 July until ~12 July, the definition of the algorithm in the package is simplified. In the process, 
this README file will not work!
</span>

# Warning!!!

The package is still under development, especially the documentation. When using
the package you might encounter bugs or errors, or you might not be able to do what you
want. In that case please write the package maintainer under his
institutional email address (requires using google).

# Installation

The package is under development. 
You can install the current development version of MoNAn from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("perblock/MoNAn")
```

or using:

``` r
# install.packages("remotes")
remotes::install_github("perblock/MoNAn")
```


# Example


In this section we outline a simple example with synthetic data stored in the MoNAn package.

```{r}
library(MoNAn)

# packages for parallel computing
library(snowfall)
```

## The Data

The example case uses synthetic data that represents mobility of 742 individuals between 17 organisations. The artificial data includes an edgelist, i.e. a list of origins (column 1) and destinations (column 2) of all individuals sorted by origin.

```{r}
mobilityEdgelist[1:10,]
```

The data further includes (artificial) characteristics of individuals, which we will call sex. However, using continuous covariates is also possible, although for some effects this would not be meaningful.

```{r}
indSex[1:10]
```

Characteristics of the organisations are the region in which they are located (binary,  northern/southern island) and the organisations' size that represents a composite measure of the number of workers, assets, and revenue.

```{r}
orgRegion
orgSize
```


## Working with the package

### Preparing the data

First, we create MoNAn data objects from the introduced data files. A necessary part is that we define and name the nodesets, i.e., we name who ("people") is mobile between what ("organisations").

```{r}
# create objects
people <- createNodeSet(nrow(mobilityEdgelist))
organisations <- createNodeSet(length(orgRegion))
transfers <- createEdgelist(mobilityEdgelist, nodeSet = c("organisations", "organisations", "people"))
sameRegion <- outer(orgRegion, orgRegion, "==") * 1
sameRegion <- createNetwork(sameRegion, nodeSet = c("organisations", "organisations"))
region <- createNodeVariable(orgRegion, nodeSet = "organisations")
size <- createNodeVariable(orgSize, nodeSet = "organisations", addSim = TRUE)
sex <- createNodeVariable(indSex, nodeSet = "people")
```

We combine the data objects into the process state, i.e., a MoNAn object that stores all information about the data that will be used in the estimation later.

```{r}
myState <- createProcessState(list(
  transfers = transfers,
  
  people = people,
  organisations = organisations,
  
  sameRegion = sameRegion,
  region = region,
  size = size,
  sex = sex
))
```


We define the dependent variable, and create a cache (a necessary object used in the estimation of the model). In case variables of the individuals in the data are included in the state (here: "sex"), they need to be explicitly mentioned in the creation of the cache under "resourceCovariates".

```{r}
myDependentVariable <- "transfers"
myCache <- createWeightedCache(myState, myDependentVariable, resourceCovariates = c("sex"))
```


### Specifying the model

The predictors in the model are called "Effects" and they are defined in a list. Each effect itself is a list that contains the effect name and the additional parameters it needs.

```{r}
# create an effects object
myEffects <- createEffectsObject(
  list(
    list("loops"),
    list("min_reciprocity"),
    list("dyadic_covariate", attribute.index = "sameRegion"),
    list("alter_covariate", attribute.index = "size"),
    list("resource_covar_to_node_covar", attribute.index = "region", resource.attribute.index = "sex"),
    list("loops_resource_covar", resource.attribute.index = "sex")
  )
)
```


### Optional: Pre-estimation

We can run a pseudo-likelihood estimation that gives a (biased) guess of the model results. We use this to get improved initial estimates, which increases the chances of model convergence in the first run of the estimation considerably. To get pseudo-likelihood estimates, we need to use functions from other libraries to estimate a multinomial logit model (e.g., "dfidx" and "mlogit")

```{r}
# create multinomial statistics object pseudo-likelihood estimation
myStatisticsFrame <- getMultinomialStatistics(myState, myCache, myEffects, myDependentVariable)

### additional script to get pseudo-likelihood estimates, requires the dfidx and mlogit package
# library(dfidx)
# library(mlogit)
# my.mlogit.dataframe <- dfidx(myStatisticsFrame,
#                           shape = "long", 
#                           choice = "choice")
# 
# colnames(my.mlogit.dataframe) <- gsub(" ", "_", colnames(my.mlogit.dataframe))
# 
# IVs <- (colnames(my.mlogit.dataframe)[2:(ncol(myStatisticsFrame)-2)])
# 
# f <- as.formula(paste("choice ~ 1 + ", paste(IVs, collapse = " + "), "| 0"))
# 
# my.mlogit.results <- mlogit(formula = eval(f), data = my.mlogit.dataframe, heterosc = F)
# 
# summary(my.mlogit.results)
#
# initEst <- my.mlogit.results$coefficients[1:length(IVs)]
```


### Estimation

Now we can estimate the model. The first two lines indicate the dependent variable, data (state), cache, and effects. The third line specifies the initial estimates, where the previously obtained pseudo-likelihood estimates can be used. The remaining lines define the algorithm (see the paper introducing the model, in particular the appendix, and the helpfiles).

Running the model takes a while (up to 10 minutes for this data with parallel computing).

```{r, eval=FALSE}
myResDN <- estimateMobilityNetwork(myDependentVariable,
  myState, myCache, myEffects,
  initialParameters = NULL,
  burnInN1 = 1500, iterationsN1 = 50, thinningN1 = 750, gainN1 = 0.1,
  burnInN2 = 7500, nsubN2 = 4, initGain = 0.2, thinningN2 = 1500,
  initialIterationsN2 = 25,
  iterationsN3 = 500, burnInN3 = 7500, thinningN3 = 3750,
  parallel = T, cpus = 4,
  allowLoops = T,
  verbose = T,
  returnDeps = T,
  multinomialProposal = F,
  fish = F
)
```


In case a pseudo-likelihood estimates have been obtained previously, this can be specified by

```{r, eval=FALSE}
myResDN <- estimateMobilityNetwork(myDependentVariable,
  myState, myCache, myEffects,
  initialParameters = initEst,
  burnInN1 = 1500, iterationsN1 = 50, thinningN1 = 750, gainN1 = 0.1,
  burnInN2 = 7500, nsubN2 = 4, initGain = 0.2, thinningN2 = 1500,
  initialIterationsN2 = 25,
  iterationsN3 = 500, burnInN3 = 7500, thinningN3 = 3750,
  parallel = T, cpus = 4,
  allowLoops = T,
  verbose = T,
  returnDeps = T,
  multinomialProposal = F,
  fish = F
)
```


Check convergence to see whether the results are reliable. In case the maximum convergence ratio is above 0.1 (or 0.2 for less precise estimates), another run is necessary.

```{r}
max(abs(myResDN$convergenceStatistics))
```

If convergence is too high, re-run estimation with previous results as starting values and check convergence:

```{r, eval=FALSE}
myResDN <- estimateMobilityNetwork(myDependentVariable,
  myState, myCache, myEffects,
  prevAns = myResDN,
  burnInN1 = 1500, iterationsN1 = 50, thinningN1 = 750, gainN1 = 0.1,
  burnInN2 = 7500, nsubN2 = 4, initGain = 0.2, thinningN2 = 3000,
  initialIterationsN2 = 40,
  iterationsN3 = 500, burnInN3 = 15000, thinningN3 = 7500,
  parallel = T, cpus = 4,
  allowLoops = T,
  verbose = T,
  returnDeps = T,
  multinomialProposal = F,
  fish = F
)
```


```{r}
# check convergence
max(abs(myResDN$convergenceStatistics))
```

In case convergence is still poor, updating the algorithm might be necessary. Otherwise, we can view the results. The first column is the estimate, the second column the standard error, and the third column the convergence ratio. All values in the final column should be below 0.1 (see above).

```{r}
myResDN
```

## Diagnostics of the estimated model

The following two diagnostics indicate the extent to which the chain mixes (i.e., whether the thinning was chosen appropriately). The autoCorrelationTest indicates the degree to which the values of the dependent variable of consecutive draws from the chain in phase 3 are correlated. Here lower values are better. Values above 0.5 are very problematic and indicate that a higher thinning is needed.

```{r}
autoCorrelationTest(myDependentVariable, myResDN)
```

The output of extractTraces indicates the correlation of statistics between subsequent draws from the chain in phase 3. The plot should show data point randomly scattered around the target line, as shown below. If patterns in the traces are discernible, a higher thinning is needed.

```{r}
traces <- extractTraces(myDependentVariable, myResDN, myEffects)

par(mfrow = c(1,2))
plot(traces)
```


## Score-tests to check model specification

Based on an estimated model, a score-type test is available that shows whether statistics representing non-included effects are well represented. If this is not the case, it is likely that including them will result in significant estimates.

```{r}
myEffects2 <- createEffectsObject(
  list(
    list("min_transitivity")
  )
)

test_ME.2 <- scoreTest(myDependentVariable, myResDN, myEffects2)
test_ME.2
```

The interpretation is that there appears to be some transitive clustering in the data that the model does not account for in its current form.


## GOF testing

Akin to ERGMs, goodness of fit testing is available to see whether auxiliary statistics are well captured by the model.

```{r}
myGofIndegree <- gofDistributionNetwork(ans = myResDN, simulations = myResDN$deps, gofFunction = getIndegree, lvls = 1:70)
plot(myGofIndegree, lvls = 1:70)

myGofTieWeight <- gofDistributionNetwork(ans = myResDN, simulations = myResDN$deps, gofFunction = getTieWeights, lvls = 1:20)
plot(myGofTieWeight, lvls = 1:20)
```

